---
title: "Cancer prediction"
author: '*Nicol&oacute; Foppa Pedretti*'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warnings = FALSE)
```

### Load libraries 

Here I load the libraries used for the analysis

```{r}
library(data.table)
library(caret)
library(randomForest)
library(parallel)
library(stringr)
```

### Load and clean data

Here I load the `clinical.csv` and `genomics.csv` datasets:

```{r}
DT = fread("clinical.csv")
genomics = fread("genomics.csv")
```

Let's start with the main task: build and evaluate a predictive model of one-year
survival after diagnosis with NSCLC. In order to answer this question I create a new variable. The main idea is create a predictive model after 12 months for dead/alive classification. My assumption is to create a new outcome value where I set to `Alive` the people that have a follow-up months longer than 12 months and the Outcome is `Dead`. That is a strong assumption because technically we don't have any date of death (which it would allow us to create this model). I map the binary like this: 

+ `Alive` &longrightarrow; 1
+ `Dead` &longrightarrow; 0

```{r}
DT[,Outcome_num := factor(ifelse(Survival.Months > 12 & Outcome == "Dead",1,0))]
```

I used `Outcome` and `Survival.Months` to create my new outcome that it will be used for the prediction. We are not using those variables anymore. Let's check the other variables 

+ `Age`: *The patient’s age (in years) at diagnosis*. Numerical variable, we don't need any transformation or cleaning. No missing data

+ `Grade`: *Tumor grade (1-4 or missing)*. 

```{r}
table(DT$Grade)
```

Using the description I can assume that the value *9* is referred to missing data. In this specific case I don't have any value `1`, but it could mean that those value are mixed in the code `9` and I cannot use any imputation on this variable because I don't have *missing at random*. Let's transform the variable in `factor`

```{r}
DT[, Grade := ifelse(Grade == 9, NA, Grade)]
DT[, Grade := factor(Grade, exclude = NULL)]
```

+ `Num.Primaries`: *Number of primary tumors*. Let's convert to factor. No missing data.

```{r}
DT[, Num.Primaries := factor(Num.Primaries, exclude = NULL)]
```

+ `T`: *Tumor Stage*. In this variable we have 62 `"UNK"` (Unknown) data and a few class with a low sample size

```{r}
table(DT$`T`)
```

I decide to re-map the (1,1a,1b) &longrightarrow; 1, (2,2a,2b) &longrightarrow; 2 and UKN &longrightarrow; `NA`  

```{r}
DT[,`T` := factor(ifelse(`T` == "UNK",NA,
                  ifelse(`T` %like% "1",1,
                  ifelse(`T` %like% "2",2,
                  ifelse(`T` %like% "3",3,4)))),exclude = NULL)]
```

+ `N`: *Number of metastasis to lymph nodes*. In this variable we have 65 `"NULL"` (NULL) data and a few class with a low sample size

```{r}
table(DT$N)
```

I collapse together classes `0`+`1` and `2`+`3` and I map `"NULL"` in `NA`

```{r}
DT[, N := factor(ifelse(N %in% c("0","1"),"01",
                 ifelse(N %in% c("2","3"),"2+",NA)),exclude = NULL)]
```

+ `M`: *Number of distant metastases*. In this variable we have 96 `"NULL"` variable and only two left unbalanced categories. I convert to factor:

```{r}
DT[, M := factor(M,exclude = NULL)]
```

+ `Radiation`: *Whether radiation took place (5) or not (0)*. No missing value and the two class are unbalanced. I map 0/5 in No/Yes and convert to a factor: 

```{r}
DT[, Radiation := factor(ifelse(Radiation == 0,"No","Yes"),exclude = NULL)]
```

+ `Stage`: *Stage at diagnosis*. This variable doesn't have missing data but have problems on sample size per category and (possibly) errors on the labels (`1B` vs. `IB`). I drop the letters (`A`,`B`) and collapse the category based on the numeric stage only:

```{r}
table(DT$Stage, useNA = "ifany")
DT[, Stage := factor(ifelse(Stage %like% "IV","IV",
                     ifelse(Stage %like% "III","III",
                     ifelse(Stage %like% "II","II","I"))))]
```

+ `Primary.Site`: *Location of primary tumor*. In this variable I don't have any missing data but I have some subclasses with few observations. I simply transform to factor the classes. Having more knowledges on lungs functionality I could collapse similar regions.

```{r}
DT[, Primary.Site := factor(Primary.Site)]
```

+ `Histology`: *The tumor histology*. In this variable I don't have any missing data, I simply transform to factor the classes. 

```{r}
DT[, Histology := factor(Histology)]
```

+ `Tumor.Size`: *Size of the tumor at diagnosis*. Continuous variable with a lot of missing data called `"NULL"`. I think that this variable could be really important. Let's check the effect with a single logit regression:

```{r}
DT$Tumor.Size = as.numeric(DT$Tumor.Size)
summary(glm(DT$Outcome_num~DT$Tumor.Size, 
            family = binomial(link = "logit")))$coeff
```

I have a lot of missing data but the coefficient is significant. I wouldn't impute this value (too many data to impute I can introduce some biases in the dataset) but I keep it

+ `Num.Mutations` and `Num.Mutated.Genes`: *The total number mutations found in the tumor and the total number of genes with mutation*. Those two variables are highly correlated because the number of mutations of each mutated gene is going from 1 to 3. Individually they don't have any significance:

```{r}
summary(glm(DT$Outcome_num~DT$Num.Mutated.Genes, 
            family = binomial(link = "logit")))$coeff

summary(glm(DT$Outcome_num~DT$Num.Mutations, 
            family = binomial(link = "logit")))$coeff
```

Let's create a ratio for the mutation. The coefficient of the ratio in logit regression is significant

```{r}
DT[, ratio_mut := ifelse(Num.Mutated.Genes > 0,
                         Num.Mutations/Num.Mutated.Genes,
                         1)]

summary(glm(DT$Outcome_num~DT$ratio_mut, 
            family = binomial(link = "logit")))$coeff
```

I keep only the `ratio_mut` variable for the analysis.

### Chi-square test

We want to test the correlation between the categorical variables and the outcome. Let's use a $\chi^2$ test to check the independency of the two variables:

```{r, message=FALSE, warning=FALSE}
cat_variable = names(DT)[c(5,6,7,8,9,10,11,12,13)]
chisq_table = lapply(cat_variable,function(i){
  data.table(Variable = i,
             p_value = round(chisq.test(DT$Outcome_num,DT[[paste0(i)]])$p.value,4))
})
chisq_table = rbindlist(chisq_table)
chisq_table
```

We reject the hypothesis of independency for all variables except for the `Num.Primaries`. We exclude the variables from the analysis.

### Genomics dataset

In order to include the genomics information into the analysis I pivot the long table of genomics to wide format to merge clinical dataset and genomic one. I also transform the variables into factors:

```{r}
genomics = dcast(genomics, ID ~ Gene, 
                 value.var = "Gene", 
                 fun.aggregate = length)
genomics[,c(names(genomics)[-1]) := lapply(.SD,function(i) factor(i,exclude = NULL)),
         .SDcols = names(genomics)[-1]]
```

### Merge the datasets

```{r}
DT = merge(DT[,.(ID,Age,Grade,`T`,N,M,Radiation,Stage,Primary.Site,
                 Histology,Tumor.Size,Outcome_num,ratio_mut)],
           genomics, by = "ID", all.x = T)
```

### Model

I use a **Random Forest** algorithm to predict the outcome. I pick this algorithm because is powerful (mean estimation across multiple classification trees), fast and it can also consider *non-linearity*. I will include all the dummy variables related to the genes that have mutations (`1` if the gene have mutation for that individual, `0` otherwise) and all the variable that showed a significant association in the analysis above. In order to perform an additional variable selection I add a random variable to my dataset. Then checking the variable importance I will exclude all the variable that will have an importance lower than my random variable:  

```{r}
DT$random_variable = rnorm(NROW(DT),0,1)
```

I have a low sample size considering the missing data, not enough to split between **train** and **test** dataset. *Random Forest algorithm* (by it's implementation) can handle that problems using the **out-of-bag error**. The out-of-bag (**OOB**) error is the average error for each calculated using predictions from the trees that do not contain in their respective bootstrap sample, in this way I can avoid the test-set and also the cross-validation. Now I run an optimization grid to look for the parameters that minimize the OOB error:

```{r, eval=FALSE}
rf_grid = data.table(expand.grid(mtry = 5:40,
                                 ntree = seq(1000,3000,250),
                                 nodesize = 1:10))
rf_grid$ID = 1:NROW(rf_grid)

n_cores = detectCores() - 2
cl = makeCluster(n_cores)
clusterEvalQ(cl, {
  library(randomForest)
  library(data.table)
  library(stringr)
})

clusterExport(cl, varlist = c("DT","rf_grid"))

rf_opt = parLapplyLB(cl,1:NROW(rf_grid), function(j){
  a = Sys.time()
  set.seed(j)
  rf = randomForest(as.factor(Outcome_num) ~ .,
                    data = DT[,-c("ID"),with=F],
                    na.action = "na.omit",
                    mtry = rf_grid$mtry[j],
                    replace = F,
                    nodesize = rf_grid$nodesize[j],
                    ntree = rf_grid$ntree[j])
  b = Sys.time()
  cat(paste0("Grid ",str_pad(j, 4, pad = "0"),
             " - ",str_pad(round(as.numeric(b-a),4),6,side = "right", pad = "0"),
             " secs - ntree: ",rf_grid$ntree[j]," - mtry: ",
             str_pad(rf_grid$mtry[j], 2, pad = "0"),
             " - nodesize: ",str_pad(rf_grid$nodesize[j], 2, pad = "0"),
             " - oob_err_rate: ",str_pad(round(rf$err.rate[rf_grid$ntree[j],1],4),6,
                                         side = "right", pad = "0"),
             "\n"),
      file = "rf_model.log",
      append = TRUE)
  return(NULL)
})
```

Checking the results from the file `rf_model.log`. Here the parameters that produced the best performance:

```{r}
rf_opt = fread("rf_model.log")
rf_opt = rf_opt[,c(2,4,8,11,14,17)]
setnames(rf_opt,names(rf_opt),c("ID","secs","ntree","mtry","nodesize","err_rate_oob"))
rf_opt[order(err_rate_oob)][1]
```

Let's explore the variable importance grid:

```{r}
set.seed(407)
rf_model = randomForest(as.factor(Outcome_num) ~ .,
                  data = DT[,-c("ID"),with=F],
                  na.action = "na.omit",
                  mtry = 15,
                  replace = F,
                  nodesize = 2,
                  ntree = 1500)
vimp = data.table(variables = rownames(varImp(rf_model)), varImp(rf_model))
vimp[order(vimp$Overall,decreasing = TRUE)]
```
In the table I can notice that only 5 variables are above my random variable. So let's try to rerun the model considering only `Tumor.Size`,`Primary.Site`,`Stage`,`Grade`,`T`. I create a new optimization grid to have the best parameters for the new model:

```{r, eval=FALSE}
rf_grid_2 = data.table(expand.grid(mtry = 1:5,
                                   ntree = seq(1000,3000,250),
                                   nodesize = 1:10))
rf_grid_2$ID = 1:NROW(rf_grid_2)

n_cores = detectCores() - 2
cl = makeCluster(n_cores)
clusterEvalQ(cl, {
  library(randomForest)
  library(data.table)
  library(stringr)
})

clusterExport(cl, varlist = c("DT","rf_grid_2"))

rf_opt = parLapplyLB(cl,1:NROW(rf_grid_2), function(j){
  a = Sys.time()
  set.seed(j)
  rf = randomForest(as.factor(Outcome_num) ~ Stage + Primary.Site + Tumor.Size + `T` + Grade,
                    data = DT[,-c("ID"),with=F],
                    na.action = "na.omit",
                    mtry = rf_grid_2$mtry[j],
                    replace = F,
                    nodesize = rf_grid_2$nodesize[j],
                    ntree = rf_grid_2$ntree[j])
  b = Sys.time()
  cat(paste0("Grid ",str_pad(j, 4, pad = "0"),
             " - ",str_pad(round(as.numeric(b-a),4),6,side = "right", pad = "0"),
             " secs - ntree: ",rf_grid_2$ntree[j]," - mtry: ",
             str_pad(rf_grid_2$mtry[j], 2, pad = "0"),
             " - nodesize: ",str_pad(rf_grid_2$nodesize[j], 2, pad = "0"),
             " - oob_err_rate: ",str_pad(round(rf$err.rate[rf_grid_2$ntree[j],1],4),6,
                                         side = "right", pad = "0"),
             "\n"),
      file = "rf_model_2.log",
      append = TRUE)
  return(NULL)
})
```

Load the file with the optimization grid results from file `rf_model_2.log`:

```{r}
rf_opt = fread("rf_model_2.log")
rf_opt = rf_opt[,c(2,4,8,11,14,17)]
setnames(rf_opt,names(rf_opt),c("ID","secs","ntree","mtry","nodesize","err_rate_oob"))
rf_opt[order(err_rate_oob)][1]
```

Analysis of the final model:

```{r}
set.seed(48)
rf_model = randomForest(as.factor(Outcome_num) ~ Stage + Primary.Site + 
                          Tumor.Size + `T` + Grade,
                  data = DT[,-c("ID"),with=F],
                  na.action = "na.omit",
                  mtry = 3,
                  replace = F,
                  nodesize = 2,
                  ntree = 1000)
rf_model
```

From the summary results we can see that the OOB estimate of  error rate is 7.14%. The model makes 24% of 1^st^ type error and 1.3% 2^nd^ type error. The performance for the 1^st^ type error is not good but it gives a more conservative prediction.  


Having more time and resources I could make a better merge between the classes, try to impute the variables that contains missing data and use the correlation/association between the variables. Having a better knowledge of the lungs I could also make assumption and/or provide a cleaner dataset for the analysis. Working with `lightgboost` or `xgboost` could provide a more refined estimations.      






